---
title: "The Frequencies of Coldplay"
subtitle: "Turning musical keys into the Hz they actually vibrate at"
author: "frequencymatch"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 120)
library(tidyverse)
```

## Why frequencies?

Every note is, underneath everything, a frequency — a number of vibrations per
second, measured in Hertz. A song written in the key of A is built around a
fundamental that oscillates at 440 Hz. Change the key, and you change the
frequency the whole piece is anchored to.

Spotify's audio analysis gives us the **key** of each track (as an integer from
0 to 11, following pitch-class notation where 0 = C, 1 = C♯, and so on). That's
usually treated as a categorical label. But a key isn't really a category — it's
a pitch, and a pitch is a frequency. So this little project does something
simple: it takes Coldplay's discography and asks a slightly unusual question.

**At what frequencies does Coldplay actually vibrate?**

## The data

The dataset covers the studio albums from *Parachutes* (2000) to
*Music of the Spheres* (2021), with each track's key, mode, tempo, energy and
valence.

```{r load}
tracks <- read_csv("data/coldplay_tracks.csv")
glimpse(tracks)
```

## From key to Hz

To turn a pitch class into a frequency, we use the standard equal-temperament
formula. Taking A4 = 440 Hz as the reference, any note *n* semitones away from A
has a frequency of:

$$ f = 440 \times 2^{\,n/12} $$

We map each pitch class (0–11) to the octave nearest the human "central" register
(the A above middle C = 440 Hz), so the numbers stay comparable across songs.

```{r key-to-hz}
# pitch-class names, 0 = C ... 11 = B
pitch_names <- c("C", "C#", "D", "D#", "E", "F",
                 "F#", "G", "G#", "A", "A#", "B")

# semitone distance of each pitch class from A (pitch class 9), within one octave
# we anchor everything to the A = 440 Hz octave
key_to_hz <- function(key) {
  # semitones from A (pitch class 9)
  semis <- key - 9
  440 * 2^(semis / 12)
}

tracks <- tracks %>%
  mutate(
    key_name = factor(pitch_names[key + 1], levels = pitch_names),
    freq_hz  = key_to_hz(key),
    mode_lab = if_else(mode == 1, "major", "minor")
  )

tracks %>%
  select(track, album, key_name, freq_hz, mode_lab) %>%
  head(10)
```

## The frequency fingerprint

Here is every Coldplay track placed on the frequency axis it's anchored to. Each
point is a song; colour marks major or minor.

```{r fingerprint, fig.height=5.5}
ggplot(tracks, aes(x = freq_hz, y = fct_reorder(album, year), colour = mode_lab)) +
  geom_jitter(height = 0.18, size = 2.6, alpha = 0.85) +
  scale_colour_manual(values = c(major = "#2dd4bf", minor = "#0e7490")) +
  labs(
    title = "The frequencies of Coldplay",
    subtitle = "Each track placed at the Hz of its fundamental key",
    x = "Fundamental frequency (Hz)", y = NULL, colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold"))
```

## Which frequencies does Coldplay return to?

Some frequencies come up again and again. Counting how often each fundamental
appears across the whole discography reveals the band's harmonic centre of
gravity.

```{r common-freqs}
freq_counts <- tracks %>%
  count(key_name, freq_hz, sort = TRUE) %>%
  mutate(freq_label = paste0(key_name, "  (", round(freq_hz), " Hz)"))

ggplot(freq_counts, aes(x = fct_reorder(freq_label, n), y = n)) +
  geom_col(fill = "#14b8a6", width = 0.7) +
  coord_flip() +
  labs(
    title = "Coldplay's most-used fundamentals",
    x = NULL, y = "Number of tracks"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## A short detour: 432 vs 440

There's a long-running idea in some corners of music that 432 Hz — rather than
the modern standard of 440 Hz — is a "more natural" tuning, closer to something
resonant in the world. It's not a scientific claim so much as a piece of
folklore, but it's a nice reminder that the *reference* frequency is a choice,
not a law of nature.

Out of curiosity, here's what Coldplay's fundamentals would look like if the
whole catalogue had been tuned to A = 432 Hz instead of 440. Everything simply
shifts down by the same small ratio.

```{r 432}
tracks <- tracks %>%
  mutate(freq_hz_432 = freq_hz * (432 / 440))

tracks %>%
  summarise(
    mean_440 = mean(freq_hz),
    mean_432 = mean(freq_hz_432),
    shift_hz = mean_440 - mean_432
  )
```

The difference is tiny — a few Hz — but it's the kind of thing you only notice
if you're listening closely.

## Has their frequency drifted over time?

Finally: as the albums went by, did the band drift toward higher or lower
fundamentals? We take the mean fundamental frequency per album and follow it
across two decades.

```{r drift}
album_freq <- tracks %>%
  group_by(album, year) %>%
  summarise(mean_freq = mean(freq_hz), .groups = "drop") %>%
  arrange(year)

ggplot(album_freq, aes(x = year, y = mean_freq)) +
  geom_line(colour = "#0e7490", linewidth = 0.9) +
  geom_point(colour = "#2dd4bf", size = 3) +
  labs(
    title = "Coldplay's average fundamental frequency, album by album",
    x = NULL, y = "Mean fundamental frequency (Hz)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Closing note

Treating a key as a frequency doesn't change the music — it just changes how you
look at it. Underneath the lyrics and the production, every one of these songs is
a set of numbers, patiently vibrating at a fixed rate, waiting to be heard.

Some frequencies find each other. Some just wait.

---

*Data: Spotify audio features (key, mode, tempo, energy, valence). Frequencies
derived from equal-temperament tuning with A4 = 440 Hz.*
